---
title: "Species Change Summary"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Overview

This document summarizes how species changed between the major assessment and the 2025 update for:

-   Long-term trend
-   Current trend
-   Range
-   Priority category

The analysis is grouped by the 2023 conservation priority .

# Input Data

```{r}
changes <- read.csv("/home/shasanko/soib_2025_rerun_bugfixed/soib_2023/01_analyses_full/results/SoIB_changes_summary.csv", header = T)

names(changes)
```

# Long-Term Trend Changes

## High Priority Species

```{r}
changes_summary_ltt_high <- changes %>%
  filter(
    Major_Priority == "High",
    Major_LongTerm != Latest_LongTerm
  ) %>%
  count(
    Major_Priority,
    Major_LongTerm,
    Latest_LongTerm
  )

changes_summary_ltt_high
```

## Moderate Priority Species

```{r}
changes_summary_ltt_mod <- changes %>%
  filter(
    Major_Priority == "Moderate",
    Major_LongTerm != Latest_LongTerm
  ) %>%
  count(
    Major_Priority,
    Major_LongTerm,
    Latest_LongTerm
  )

changes_summary_ltt_mod
```


# Current Trend Changes

## High Priority Species

```{r}
changes_summary_cat_high <- changes %>%
  filter(
    Major_Priority == "High",
    Major_Current != Latest_Current) %>%
  count(
    Major_Priority,
    Major_Current,
    Latest_Current
  )

changes_summary_cat_high
```

## Moderate Priority Species

```{r}
changes_summary_cat_mod <- changes %>%
  filter(
    Major_Priority == "Moderate",
    Major_Current != Latest_Current) %>%
  count(
    Major_Priority,
    Major_Current,
    Latest_Current
  )

changes_summary_cat_mod
```

# Range Changes (all priorities combined)

```{r}
changes_summary_range <- changes %>%
  filter(Major_Range != Latest_Range) %>%
  count(
    Major_Priority,
    Major_Range,
    Latest_Range
  )

changes_summary_range %>%
  arrange(Major_Priority, desc(n))
```


# Total Number of Species With Changes

```{r}
overall_summary <- changes %>%
  summarise(
    LongTerm_Changes = sum(Major_LongTerm != Latest_LongTerm, na.rm = TRUE),
    Current_Changes  = sum(Major_Current != Latest_Current, na.rm = TRUE),
    Range_Changes    = sum(Major_Range != Latest_Range, na.rm = TRUE),
    Priority_Changes = sum(Major_Priority != Latest_Priority, na.rm = TRUE)
  )

overall_summary
```

# Summary by Major Priority

```{r}
summary_by_priority <- changes %>%
  group_by(Major_Priority) %>%
  summarise(
    Total_Species = n(),

    LongTerm_Changes =
      sum(Major_LongTerm != Latest_LongTerm, na.rm = TRUE),

    Current_Changes =
      sum(Major_Current != Latest_Current, na.rm = TRUE),

    Range_Changes =
      sum(Major_Range != Latest_Range, na.rm = TRUE),

    Priority_Changes =
      sum(Major_Priority != Latest_Priority, na.rm = TRUE)
  )

summary_by_priority
```

